---
title: "Comparing page load times"
author: "Dominik Strohmeier, Mozilla"
date: "11/11/2017"
output:
  html_document:
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

This notebook analysis data that was collected by [Disconnect](https://disconnect.me/) and [Mozilla](https://www.mozilla.org/en-US/) to compare page load speed of different browsers.

Browsers under test:

* Google Chrome v62
* Google Chrome in Incognito mode
* Firefox Quantum (v57)
* Firefox Quantum (v57) in Private Browsing mode with Tracking Protection

```{r data, message=FALSE, warning=FALSE}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(scales, warn.conflicts = FALSE)
library(data.table, warn.conflicts = FALSE)

# colorblind-friendly colors
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# read data from results csv file. Reading timestamps as character to keep it readible.
tp_data <- fread("pageloadstudy.csv", colClasses = list(character=3:25))

# replace strings for different Browsers
tp_data <- tp_data[Browser == "chrome_normal", Browser := "Chrome"]
tp_data <- tp_data[Browser == "chrome_private", Browser := "Chrome Incognito"]
tp_data <- tp_data[Browser == "firefox_normal", Browser := "Firefox Quantum"]
tp_data <- tp_data[Browser == "firefox_private", Browser := "Firefox Quantum Private Browsing"]

```

In our study, we collected [performance.timing](https://developer.mozilla.org/en-US/docs/Web/API/PerformanceTiming) data for 200 news websites that were loaded 10 times each in every browser. Here is a sample of the data as example:

```{r preview, message=FALSE, warning=FALSE}
# create difference between two timestamps
# we are interested in the difference between each of the timestamps and navigationStart
perf_timings <- tp_data %>% 
  mutate_at(.vars = vars(-Domain, -Browser, -`Load Time`), .funs = funs(diff = as.numeric(.)-as.numeric(navigationStart))) %>% 
  select(Domain, Browser, ends_with("diff")) %>% 
  mutate_if(is.numeric, funs(replace(., .<0, NA))) %>% 
  select_if(~sum(!is.na(.)) > 0)

names(perf_timings) <- gsub("_diff", "", names(perf_timings))

#display sample of data_perf in notebook
sample_n(perf_timings, 3)
```


## Comparing means per browser

To analyze the data, let's look for at the mean page load time for each browser and compare differences.

```{r means}

by_browser <- perf_timings %>% 
  group_by(Browser) %>% 
  summarize(mean_page_load_time = mean(loadEventEnd))

by_browser

ggplot(data = perf_timings, aes(Browser, loadEventEnd)) +
  geom_boxplot() +
  stat_summary(fun.y=mean, colour="#E69F00", geom="point", shape=16, size=5,show.legend = F) + scale_y_log10( breaks = c(100, 500, 1000, 5000, 10000)) + 
  theme_minimal() +
  theme(text = element_text(size=16)) +
  scale_x_discrete(labels=c("Chrome" = "Chrome", "Chrome Incognito" = "Chrome\nIncognito", "Firefox Quantum" = "Firefox\nQuantum", "Firefox Quantum Private Browsing" = "Firefox\nQuantum\nPrivate\nBrowsing")) + 
  labs(y = "time to loadEventEnd in ms")

```

The difference between Chrome's Incognito mode and Firefox Quantum Private Browsing is **`r round(by_browser$means[by_browser$Browser == "Chrome"] / by_browser$means[by_browser$Browser == "Firefox Quantum Private Browsing"], 1)`x**. It is also visible that there is no difference between Chrome's normal and Incognito mode. This shows that the differences between Firefox Quantum and its Private Browser option, which is similar to Chrome's Incognito mode + Tracking Protection, come from Tracking Protection.

## Putting simulated measurements into context
One important final step when doing automated performance measurements is to find related results that can be used to put your measurements into context. 

[SOASTA](www.soasta.com) published results about predicting user bounce rate based on performance timings that they measured in the wild. Average session load time was one of the best performing predictors in [their study](https://www.slideshare.net/tammyeverts/using-machine-learning-to-determine-drivers-of-bounce-and-conversion-66319405).

Their results show that an average session load time of 6 seconds already leads to a 70% bounce rate. Let's look at the share of pages in our data that has a load time longer than 6 seconds and compare across browsers.

```{r bounce}
long_loading <- perf_timings %>% 
  filter(grepl("Incog|Private", Browser)) %>% 
  mutate(islonger = ifelse(loadEventEnd > 6000, 1, 0)) %>% 
  group_by(Browser) %>% 
  summarize(slow = sum(islonger), fast = n() - slow) %>% 
  gather(page_load_time, n, slow:fast) %>% 
  mutate(perc = (n / 2000)*100)

long_loading$page_load_time <- factor(long_loading$page_load_time, levels=c("fast", "slow"), labels=c("within 6 sec", "longer than 6 sec"))

ggplot(long_loading, aes(x = Browser, y = perc, fill = page_load_time, label = perc)) + 
  geom_bar(stat = "identity", position = "fill", width = 0.5) + 
  scale_y_continuous(labels=percent) + 
  theme_minimal() +
  theme(text = element_text(size=16)) +
  scale_x_discrete(labels=c("Chrome" = "Chrome", "Chrome Incognito" = "Chrome\nIncognito", "Firefox Quantum" = "Firefox\nQuantum", "Firefox Quantum Private Browsing" = "Firefox\nQuantum\nPrivate\nBrowsing")) +
  labs(y = "Share of pages loaded in %", color = "Page load time", fill = "Page load time") + 
  scale_fill_manual(values=cbPalette) +
  geom_text(size = 4,  position = position_fill(vjust = 0.5))
```

While only about 5.5% of page loads take longer than 6 seconds for Firefox Quantum, it's about 31% of all pages for Google Chrome.

## Looking at where the time differences occur
A last, interesting aspect to look into the data is to look at the performance timing timeline and check where the differences between browsers occur. Let's only look at Chrome Incognito and Firefox Quantum Private Browsing again.

*performance.timing* gives you events along the page load process. Let's print these events in order of appearance during page load and look at differences between browsers, looking at [newsweek.com](http://newsweek.com) as an example:

```{r, message=FALSE, warning=FALSE}
# order of appearance of events during page load
standard = c("navigationStart", "redirectStart", "fetchStart", "domainLookupStart", "domainLookupEnd", "connectStart", "connectEnd", "responseStart", "responseEnd", "domLoading", "domInteractive", "domContentLoadedEventStart", "domContentLoadedEventEnd", "domComplete", "loadEventStart", "loadEventEnd")

perf_timings %>%
  select(-secureConnectionStart, -redirectStart, -redirectEnd, -unloadEventEnd, -unloadEventStart) %>%
  filter(Domain == "http://Newsweek.com" & grepl("Incog|Private", Browser)) %>%
  gather(metric, time, -Domain, -Browser) %>%
  mutate(metric = factor(metric, levels = standard)) %>%
  na.omit() %>%
  ggplot(aes(metric, time, fill = Browser)) + geom_boxplot() +
  scale_fill_manual(values=cbPalette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size=14)) +
  labs(y = "time after navigationStart in ms", x = "performance.timing events", title = "newsweek.com")

```

It becomes visible that the main differences occur towards the end of loading process. Work to create the DOM is doing similarly fast in both browsers, but Chrome is waiting for content significantly longer than Firefox as main differences start to occur with domComplete, i.e. the moment in time when the parser finished its work on the main document. This underlines again that Firefox's Tracking Protection used in Private Browsing blocks slow, very likely third party content from being loaded.
